Tree growth, mortality, and climatic water deficit in west-coast states, USA

### 2025-12-16 Manuscript.Rmd
Next revision version (first complete revision after B.Frank's edits).  The permutation analysis has been removed and a new figure of mean species initial/delta CWD values and a new map of negative initial/delta CWD values have been added.

x <- data.2use %>% arrange(get(col.interest))
x2 <- x[place2, ]
x2.range <- paste0("(5th percentile = ", x2$q05, "mm, 95th percentile = ", x2$q95, "mm)")
x2.range.full <- paste0("(5th-95th percentile range = ", x2$range05.95, "mm, 5th percentile = ", x2$q05, "mm, 95th percentile = ", x2$q95, "mm)")
x2.mean.full <- paste0("(mean = ", x2$Mean, "mm, 5th percentile = ", x2$q05, "mm, 95th percentile = ", x2$q95, "mm)")

@@

# 
fig_nums <- captioner::captioner(prefix = "")
tab_nums <- captioner::captioner(prefix = "Table")

@@

Western North American forests are experiencing accelerating tree mortality rates in response to intensifying drought and water stress driven by climate change.  Tree mortality and growth reflect cumulative effects of direct physiological stress (xylem cavitation, carbon starvation), and indirect climate-mediated impacts (wildfire, insects, pathogens).  Relating design-based estimates for mortality and growth to a climate variable can be challenging, as directly assessing mortality and growth requires the visitation of plots at two different points in time.  Plots can differ both in their initial climate variable value and the change in that value between visits, necessitating estimation domains that jointly consider initial conditions and temporal change. We developed an analysis framework that partitions plots into domains defined by their initial-visit climatic water deficit (CWD) value and the change in the CWD value between visits.  We applied this approach to `r comma(analysis.stats$plot.num)` National Forest Inventory plots spanning California, Oregon, and Washington, tracking `r comma(analysis.stats$tree.num)` trees from `r n.spp` species over decade-long intervals.  We estimated decadal growth and mortality overall, by state, and across six CWD domains per species, along with associated measures of uncertainty.  Species responses were highly variable, with no universal pattern of increased mortality or reduced growth in domains experiencing elevated water stress. Several species exhibited overall mortality rates exceeding 0.20 (i.e., 20% decadal mortality), with estimates of 0.34 for *Picea engelmannii* in Washington, 0.27 for *Acer macrophyllum* in California, and 0.37 for *Pinus contorta* in Washington. Domain-specific patterns aligned with known regional disturbances (bark beetles, drought, fire) for multiple species. Our study contributes a template for monitoring forest vital rates as ecosystem-level indicators of climate change response, and our species-specific findings identify or validate populations warranting immediate conservation assessment that can inform forest management strategies.

@@

Climate change as manifest through increases or intensification in drought and aridity can affect tree mortality directly and indirectly.  Direct impacts include physiological damage while indirect causes include increasing wildfire, insect, and disease outbreak frequency and severity.  Hammond et al. (2022) found globally that the intensity of drought and heat events are directly associated with widespread forest die-off events.  Wildfire frequency and area burned in the western United States has increased in concert with the longer fire seasons, hotter summer temperatures, and earlier snowmelt (Abatzoglou and Williams 2016, Westerlin 2016, Juang et al. 2022).  There has also been a regional increase in the area of bark beetle (subfamily *Scolytinae* spp.) outbreaks (Meddens et al. 2012, Hicke et al. 2016).  Many of these indirect climate-influenced mortality causes may be attributable to the physical condition of trees following abnormal or prolonged exposure to water stress.

@@

Data from the U.S. National Forest Inventory, collected by the US Forest Service Forest Inventory and Analysis Program (FIA), is well suited for evaluating regional species mortality and growth. Since 1999 the FIA program has gathered a variety of forestry data at plots nationwide using a spatially balanced sampling design (McRoberts 2005, Reams et al. 2005, Woudenberg et al. 2010).  The sampling design allows for approximately unbiased design-based inference of population characteristics across large geographical regions.  Data from plot revisits and individual tree data permit direct estimation of change features such as mortality and growth estimation.

@@

Our analysis relies on FIA data for forestland in California, Oregon, and Washington, USA.  Forestland is defined as an area of at least 4,050 m$^2$ (i.e., one acre) that is at least 10 percent stocked or potentially stocked with tree species, excluding urban and agricultural land uses (Reams et al. 2005).  We constrained our analysis to only include remeasurements for areas considered forestland during both visits.

In the coterminous United States, the FIA program uses a spatially balanced, quasi-systematic sampling design. Plots are randomly located and produce sampling intensities of approximately 2,400 hectares per plot, with some denser areas such as those located on national forest system lands. In the western United States, each plot is assigned to one of ten panels, and approximately one panel is measured each year in perpetuity, yielding ten year remeasurement intervals on each plot, barring exceptions owing to optional temporal intensifications (Thompson 2015). In this study, we use only those plots that are part of the standard sampling intensity, and have been remeasured at an interval of ten years. Each FIA plot consists, in part, of four 7.32 m radius subplots where all trees with stem diameters greater than or equal to 12.7 cm (for most species measured at 1.37 m above the ground) were tallied with additional trees sampled in an annulus around each subplot (macroplot), with a total radius 18 m from the center point and a total area of 1012 m$^2$.  We did not include data from microplots, where seedlings and saplings are measured, as the analysis was not intended to estimate species recruitment but rather to follow the trajectory of established trees. Reams et al. (2005) and Bechtold and Scott (2005) provide plot design details and measurement protocols.

Our analysis depends on examining the fate and growth of trees measured during two sequential visits (referred to as the first and second plot visit), ideally from a complete sample panel remeasure. Tree growth is the observed increase in marked tree bole diameter between visits (cm$^2$) while mortality is captured from the change in tree condition. We obtained our FIA data from the USDA FIA DataMart web portal (https://research.fs.usda.gov/products/dataandtools/fia-datamart; downloaded May 5, 2025).  The database included plot data sampled from 2001 to 2023.

@@

We coarsely categorized tree species by several taxonomic and ecological characteristics, such as tree type (angiosperm [hardwood] vs. conifer), drought tolerance, fire adaption, elevation, shade tolerance, and mesic and riparian specialization.  We obtained the category information from https://plants.usda.gov/ (accessed 12/12/2025) and Burns and Honkala (1990a, 1990b).  Categorization was done to aid the discussion of results; some species could fall under more than one category (e.g., we classified *Picea engelmannii* as a high elevation conifer [HEC] while it could have been classified as a shade-tolerant conifer [STC]).  Our categories with abbreviations also include fire-adapted conifers (FAC), drought-tolerant conifers (DTC), drought-tolerant hardwoods (DTH) and riparian and mesic hardwoods (RMH).  These abbreviations are used in Tables `r tab_nums("tbl_growth", display = "num")` and `r tab_nums("tbl_mort", display = "num")`.

## Analysis
We limited the number of species examined to those with more than 400 occupied plots across the states of Oregon, Washington, and California.  We obtained point estimates of species growth and mortality using ratio estimators (Eqn 4.16 in Scott et al. 2005).  We expanded the number of sampled trees in each plot by the area measured (subplot or macroplot area) adjusted by the proportion of the plot area available for measurement and the number of years between visits. The numerator for species-specific mortality is the sum of stratum-weighted means of the number of trees that died between visits per plot.  The denominator for mortality is the sum of stratum-weighted means of all live trees per plot of a given species at the initial visit. Thus, the estimator describes the proportion of trees that transitioned from live to dead between remeasurements.

The numerator of species-specific growth estimates is the sum of stratum-weighted means of the increase in basal area (cm$^2$) between visits per plot while the denominator is the same as was used for the mortality estimate. This produces a point estimate for a species’ mean growth on an individual tree basis.  We standardized mortality and growth values to a decadal time period.

We report species mortality and growth estimates for all plots across and within Oregon, Washington, and California. Separately, we investigate climatic effects on tree mortality and growth by partitioning trees into subpopulations (domains).  For individual species we estimate domain means and measures of uncertainty for both mortality and growth. In this study we use six domains based on CWD values. We created the domains based on initial CWD plot values by partitioning the values into three species-specific quantiles: values below the first quartile (Low), between the first and third quartiles (Medium), and above the third quartile (High; Figure `r fig_nums("fig_domains", display = "num")`C).  We defined the two domains for delta CWD plot values for all species as those above a 5mm CWD change threshold, which we interpret as increasing in CWD (Increasing), and those below the threshold (Stable). The 5mm threshold was selected because species' delta CWD value distributions did not differ greatly and because 5mm is approximately the mean delta CWD value across species (Table S2).  The Stable domains may include some negative delta CWD values (i.e., mean CWD was larger prior to the initial visit than between visits), however these instances were rare and of small magnitude. The domain designations IL, IM, and IH correspond to Increasing Low, Increasing Medium, and Increasing High. The Stable domains are similarly labeled (SL, SM, SH). These six within-domain estimates allowed us to establish mortality and growth rates for trees that experienced similar climatic conditions for a given species. We calculated estimates for all CWD-based domains with 10 or more plots.

In addition to within-domain estimates, we examined contrasts among growth (Figure `r fig_nums("fig_diffgrow", display = "num")`) and mortality (Figure `r fig_nums("fig_diffmort", display = "num")`) across CWD domains. This allows us to establish if trees between two CWD domains are experiencing meaningful differences in growth and mortality. For example, this allows us to ask whether trees in historically dry sites (High initial CWD) respond differently to increased water stress (Increasing, or above threshold delta CWD) compared to trees in mesic sites (Low initial CWD) experiencing similar increases, among other comparisons.

We obtained empirical means and 95% confidence intervals via a nonparametric bootstrap. For each of the 1000 bootstrap iterations we sampled FIA plots with replacement and obtained point estimates for each domain, yielding a maximum of `r n.spp * 6 ` in-domain point estimates for each iteration.  For each domain we calculated a mean, upper, and lower 95% CI with the 2.5% and 97.5% quantiles from the  bootstrap iterations.  This approach allows for asymmetrical CIs around means.  We estimated contrasts between pairs of CWD domains by subtracting within-iteration domain point estimates.

Our  contrast analysis approach resulted in hundreds of comparisons. Given this large number of tests, we caution readers to interpret individual significant results (i.e., contrast 95% CIs that do not including zero) cautiously, as some proportion likely represent Type-I errors and occurred by chance.  We did not apply a multiple comparisons correction as our analysis is inherently exploratory, aimed at identifying patterns within and across species rather than testing a specific *a priori* hypothesis.  We did not want to apply corrections because we did not want to inflate the Type-II error rate and obscure real relationships in favor of reducing false positives.

## Dashboard
We developed a data dashboard (https://tinyurl.com/FIAgrowmort) to enable ourselves and readers to examine individual species domain estimates, the distribution of initial and delta CWD values, and the geographical distribution of plots and plot domains.  All analysis files, manuscript markdowns, dashboard code, and ancillary data sources are available at the GitHub repository https://github.com/jeremygroom/Tree_Growth_Death.

@@

Delta CWD increased at `r analysis.stats$cwd.gt0.pct`% of plots. The increases occurred in all forested regions of the three states (Figure `r fig_nums("fig_cwdmap", display = "num")`C).  Plots where CWD decreased (indicated by negative delta CWD values) are presented alone in Figure S1.  For Washington and Oregon negative CWD values were almost entirely found in the Coast Ranges and the mesic west slope of the Cascades Range crest.  The Blue Mountains and Northern and northeastern Washington lacked negative delta CWD plots entirely. In California negative delta CWD plots with values < 1mm were primarily found in the northern Sierra Nevada and in the southern portions of the state. Overall, initial and delta CWD values were weakly and positively linearly associated (correlation = `r CWD.lin.cor`).The two variables were better modeled with a 3rd order polynomial although the R^2^ value still indicated that the variance was not well explained by the model ($\Delta AIC$ =  `r CWD.dAIC`, R^2^ = `r round(CWD.rels$lm2$r.squared, 2)`, Figure S2, see model fits in Tables S1 and S2), implying that other variables likely affect the relationship between initial and delta CWD. For most species delta CWD neither linearly related to nor monotonically increased with initial CWD values (Figure S3; see dashboard). For instance, the relationship between the initial and delta CWD values for *Calocedrus decurrens* may be described overall with a concave curve, with some of the most negative delta CWD values occurring in the High domain (along with some of largest delta CWD values; Figures `r fig_nums("fig_domains", display = "num")`C).

@@

Initial CWD means and ranges differed among tree species (Table S3).  *`r init.mean.sm$spp.out`* had the smallest mean value for initial CWD (`r init.mean.sm$mean.alone` mm) while *`r init.range.sm$spp.out`* had both the largest mean value (`r init.mean.lg$mean.alone` mm) and smallest range `r init.range.sm$spp.range`.  *`r init.range.lg$spp.out`* occupied the plots with the largest range of initial CWD values `r init.range.lg$spp.range` and was the most widely distributed of our species (dashboard).

@@

Overall growth rates differed by species but in some instances coincided with species’ ecological traits.  For instance, the drought-tolerant hardwoods (*Arbutus menziesii*, *Notholithocarpus densiflorus*, *Quercus chrysolepis*, and *Quercus kelloggii*) exhibited the lowest growth rate, with all mean decadal basal area increases < 75 cm^2^ (Table 1).  *Quercus chrysolepis* had the least mean growth at 40.5 cm^2^/decade.  Riparian and mesic hardwoods like *Acer macrophyllum* and *Alnus rubra* on average grew by 145 cm^2^/decade, although *Acer macrophyllum* growth varied by state; its mean growth in California was 52.3 cm^2^/decade, less than half of its growth rate in Oregon and almost four times less than its growth rate in Washington.  High-elevation conifers like *Abies concolor* and *Picea engelmannii* would both have grown at similar rates of approximately 170-190 cm^2^/decade except that the Washington rate for *Picea engelmannii* was only 125.4 cm/decade.  The other high-elevation conifers, *Abies lasiocarpa*,  and *Tsuga mertensiana*, grew more slowly at almost 70 cm^2^/decade.

@@

Individual species growth and mortality domain patterns could differ (e.g., Figures `r fig_nums("fig_domains", display = "num")`A and `r fig_nums("fig_domains", display = "num")`B). Growth domain contrasts (Figure `r fig_nums("fig_diffgrow", display = "num")`, `r t.ref("tbl_summ_grow_mort")`) did not appear to result in substantially more significant  negative or positive estimates, indicating that species overall did not exhibit more or less growth in the Increasing, Low, or High domains. The figure column “Increasing vs. Stable” displays the mean point estimate and 95% CI for contrast between corresponding Increasing and Stable domains (e.g., for *Abies lasiocarpa*, subtracting the estimates for columns SL, SM, and SH from IL, IM, IH, respectively; Figure `r fig_nums("fig_domains", display = "num")`A). Negative estimates for the “Increasing vs. Stable” in Figures `r fig_nums("fig_diffgrow", display = "num")` and `r fig_nums("fig_diffmort", display = "num")` indicate that the 10-year growth or mortality rate was greater for a species in plots below the change estimates. `r t.ref("tbl_summ_grow_mort")` summarizes the number of domain differences in each of the columns in Figures `r fig_nums("fig_diffgrow", display = "num")` and `r fig_nums("fig_diffmort", display = "num")` where the 95% CI does not include zero as well as the number of contrasts per figure column.

We observed varying growth responses across the CWD domains (Figure `r fig_nums("fig_diffgrow", display = "num")`). Depending on the species, growth may have been susceptible (negatively or positively) to a change in CWD and/or differed by initial CWD values. We consider a lower initial CWD value to indicate a more favorable condition for plant growth, and indeed *Abies amabilis*, *Abies grandis*, *Calocedrus decurrens*, *Pinus ponderosa*, *Pseudotsuga menziesii*, *Tsuga mertensiana*, *Acer macrophyllum*, and *Alnus rubra* exhibited greater growth in the Low domains than the Medium and/or High domains. This was exemplified by *Acer macrophyllum* and *Abies grandis* with a mean difference of `r fig.val.fcn("g", "X312", "IH - IL", 0)` cm^2^/decade, IH-IL and `r fig.val.fcn("g", "X312", "SH - SL", 0)` cm^2^/decade, SH-SL, respectively while contrasts for species like *Tsuga mertensiana* and *Quercus chrysolepis* were less (-0.20 for IM-IL and SM-SL, and `r fig.val.fcn("g", "X805", "SH - SM", 0)` for SH-SM, respectively).  Except for *Alnus rubra*, these results predominantly occurred in both the Stable domain (2 spp.) or the Increasing and Stable domains (5 spp.). However, this finding is not uniform for all species, and we found that *Thuja plicata* had greater growth in the Increasing (`r fig.val.fcn("g", "X242", "IH - IM", 0)` cm^2^/decade for IH-IM and `r fig.val.fcn("g", "X242", "IM - IL", 0)` cm^2^/decade for IM-IL) and Stable (`r fig.val.fcn("g", "X242", "SH - SM", 0)` cm^2^/decade for SH-SM and `r fig.val.fcn("g", "X242", "SM - SL", 0)` cm^2^/decade for SM-SL) Medium domains than either High or Low domains.  *Abies concolor*, *Larix occidentalis*, *Picea engelmannii*, *Pinus contorta*, *Pinus lambertiana*, *Arbutus menziesii*, and *Quercus kelloggii* had greater growth in the more water stressed High domains than the Medium and/or Low domains, with some of the greatest differences for *Pinus lambertiana* (`r fig.val.fcn("g", "X117", "IH - IL", 0)` cm^2^/decade for IH-IL) and *Picea engelmannii* (`r fig.val.fcn("g", "X93", "SH - SL", 0)` cm^2^/decade for SH-SL).  Except for *Larix occidentalis*, these results predominantly occurred in both the Increasing domain (3 spp.) or the Increasing and Stable domains (3 spp.). Other species exhibited no significant findings (4 spp.) or displayed other result patterns (3 spp.).

@@

Contrast results serve to highlight differences in mortality and growth rates for different subpopulations of trees undergoing different climatic experiences. Regional analyses detected several species with relatively high mortality rates including *Acer macrophyllum*, *Picea engelmannii*, *Abies lasiocarpa*, *Pinus contorta*, and *Alnus rubra*.  We do not believe our results are concerning for *Alnus rubra*.  Our estimates for this short-lived tree species align with reported background mortality rates (e.g., Brown and Courtin 2018).

>>>>

x <- data.2use %>% arrange(get(col.interest))
x2 <- x[place2, ]
x2.range <- paste0("(5th percentile = ", x2$q05, " mm, 95th percentile = ", x2$q95, " mm)")
x2.range.full <- paste0("(5th-95th percentile range = ", x2$range05.95, " mm, 5th percentile = ", x2$q05, " mm, 95th percentile = ", x2$q95, " mm)")
x2.mean.full <- paste0("(mean = ", x2$Mean, " mm, 5th percentile = ", x2$q05, " mm, 95th percentile = ", x2$q95, " mm)")

@@

# 
fig_nums <- captioner::captioner()
tab_nums <- captioner::captioner(prefix = "Table")

@@

Western North American forests are experiencing accelerating tree mortality rates in response to intensifying drought and water stress driven by climate change.  Tree growth and mortality reflect cumulative effects of direct physiological stress (xylem cavitation, carbon starvation), and indirect climate-mediated impacts (wildfire, insects, pathogens).  Relating design-based estimates for mortality and growth to a climate variable can be challenging, as directly assessing mortality and growth requires the visitation of plots at two different points in time.  Plots can differ both in their initial climate variable value and the change in that value between visits, necessitating estimation domains that jointly consider initial conditions and temporal change. We developed an analysis framework that partitions plots into domains defined by their initial-visit climatic water deficit (CWD) value and the change in the CWD value between visits.  We applied this approach to `r comma(analysis.stats$plot.num)` National Forest Inventory plots spanning California, Oregon, and Washington, tracking `r comma(analysis.stats$tree.num)` trees from `r n.spp` species over decade-long intervals.  We estimated decadal growth and mortality across three states, by state, and for six CWD domains per species, along with associated measures of uncertainty.  Species responses were highly variable, with no universal pattern of increased mortality or reduced growth in domains experiencing elevated water stress. Several species exhibited overall mortality rates exceeding 0.20 (i.e., 20% decadal mortality), with estimates of 0.34 for *Picea engelmannii* in Washington, 0.27 for *Acer macrophyllum* in California, and 0.37 for *Pinus contorta* in Washington. Domain-specific patterns aligned with known regional disturbances (bark beetles, drought, fire) for multiple species. Our study contributes a template for monitoring forest vital rates as ecosystem-level indicators of climate change response, and our species-specific findings identify or validate populations warranting immediate conservation assessment that can inform forest management strategies.

@@

Climate change as manifest through increases or intensification in drought and aridity can affect tree mortality directly and indirectly.  Direct impacts include physiological damage while indirect causes include increasing wildfire, insect, and disease outbreak frequency and severity.  Hammond et al. (2022) found globally that the intensity of drought and heat events are directly associated with widespread forest die-off events.  Wildfire frequency and area burned in the western United States has increased in concert with the longer fire seasons, hotter summer temperatures, and earlier snowmelt (Abatzoglou and Williams 2016, Westerlin 2016, Juang et al. 2022). There have also been regional increases in the area of bark beetle (subfamily *Scolytinae* spp.) outbreaks (Meddens et al. 2012, Hicke et al. 2016).  Many of these indirect climate-influenced mortality causes may be attributable to the physical condition of trees following abnormal or prolonged exposure to water stress.

@@

Data from the U.S. National Forest Inventory, collected by the US Forest Service Forest Inventory and Analysis Program (FIA), is well suited for evaluating regional species growth and mortality. Since 1999 the FIA program has gathered a variety of forestry data at plots nationwide using a spatially balanced sampling design (McRoberts 2005, Reams et al. 2005, Woudenberg et al. 2010).  The sampling design allows for approximately unbiased design-based inference of population characteristics across large geographical regions.  Data from plot revisits and individual tree data permit direct estimation of change features such as mortality and growth estimation.

@@

Our analysis relies on FIA data for forestland in California, Oregon, and Washington, USA.  Forestland is defined as an area of at least 4,050 m^2^ (i.e., one acre) that is at least 10 percent stocked or potentially stocked with tree species, excluding urban and agricultural land uses (Reams et al. 2005).  We constrained our analysis to only include remeasurements for areas considered forestland during both visits.

In the coterminous United States the FIA program uses a spatially balanced, quasi-systematic sampling design. Plots are randomly located and produce sampling intensities of approximately 2,400 hectares per plot, with some denser areas such as those located on national forest system lands. In the western United States, each plot is assigned to one of ten panels, and approximately one panel is measured each year in perpetuity, yielding ten year remeasurement intervals on each plot, barring exceptions owing to optional temporal intensifications (Thompson 2015). In this study, we use only those plots that are part of the standard sampling intensity and have been remeasured at an interval of ten years. Each FIA plot consists, in part, of four 7.32 m radius subplots where all trees with stem diameters greater than or equal to 12.7 cm (for most species measured at 1.37 m above the ground) were tallied with additional trees sampled in an annulus around each subplot (macroplot), with a total radius 18 m from the center point and a total area of 1012 m^2^.  We did not include data from microplots, where seedlings and saplings are measured, as the analysis was not intended to estimate species recruitment but rather to follow the trajectory of established trees. Reams et al. (2005) and Bechtold and Scott (2005) provide plot design details and measurement protocols.

Our analysis depends on examining the fate and growth of trees measured during two sequential visits (referred to as the first and second plot visit), ideally from a complete sample panel remeasure. Tree growth is the observed increase in marked tree bole diameter between visits (cm^2^) while mortality is captured from the change in tree condition. We obtained our FIA data from the USDA FIA DataMart web portal (https://research.fs.usda.gov/products/dataandtools/fia-datamart; downloaded May 5, 2025).  The database included plot data sampled from 2001 to 2023.

@@

We coarsely categorized tree species by several taxonomic and ecological characteristics, such as tree type (angiosperm [hardwood] vs. conifer), drought tolerance, fire adaption, elevation, shade tolerance, and mesic and riparian specialization.  We obtained the category information from https://plants.usda.gov/ (accessed 12/12/2025) and Burns and Honkala (1990a, 1990b).  Categorization was done to aid the presentation of results; some species could fall under more than one category (e.g., we classified *Picea engelmannii* as a high elevation conifer [HEC] while it could have been classified as a shade-tolerant conifer [STC]).  Our categories with abbreviations also include fire-adapted conifers (FAC), drought-tolerant conifers (DTC), drought-tolerant hardwoods (DTH) and riparian and mesic hardwoods (RMH).  These abbreviations are used in Tables `r tab_nums("tbl_growth", display = "num")` and `r tab_nums("tbl_mort", display = "num")`.

## Analysis
We limited the number of species examined to those with more than 400 occupied plots across the states of Oregon, Washington, and California.  We obtained point estimates of species growth and mortality using ratio estimators (Eqn 4.16 in Scott et al. 2005).  We expanded the number of sampled trees in each plot by the area measured (subplot or macroplot area) adjusted by the proportion of the plot area available for measurement and the number of years between visits.  The mortality estimator describes the proportion of trees that transitioned from live to dead between remeasurements. For a given species the numerator is the sum of stratum-weighted means, where each mean is the sum of trees that died between visits within each stratum divided by the number of plots within the stratum.  The denominator for mortality is also a sum of stratum-weighted means, except each mean is the sum of all live trees within a stratum at the initial visit divided by the number of plots within the stratum.

The numerator of species-specific growth estimates is a sum of stratum-weighted means, where each mean is the sum of tree basal area increase (cm^2^) between visits within the stratum divided by the number of plots in the stratum while the denominator is the same as was used for the mortality estimate. This produces a point estimate for a species’ mean growth on an individual tree basis.  We standardized mortality and growth values to a decadal time period.

We report species mortality and growth estimates for all plots across ("overall") and within Oregon, Washington, and California. Separately, we investigate climatic effects on tree mortality and growth by partitioning trees into subpopulations (domains).  For individual species we estimate domain means and measures of uncertainty for both mortality and growth. In this study we use six domains based on CWD values. We created the domains based on initial CWD plot values by partitioning the values into three species-specific quantiles: values below the first quartile (Low), between the first and third quartiles (Medium), and above the third quartile (High; Figure `r fig_nums("fig_domains", display = "num")`C).  We defined the two domains for delta CWD plot values for all species as those above a 5 mm CWD change threshold, which we interpret as increasing in CWD (Increasing), and those below the threshold (Stable). The 5 mm threshold was applied across species because species' delta CWD value distributions did not differ greatly and because 5 mm is approximately the grand mean delta CWD value (Table S2).  The Stable domains may include some negative delta CWD values (i.e., the mean CWD value was larger prior to the initial visit than between visits); however, these instances were rare and of small magnitude. The domain designations IL, IM, and IH correspond to Increasing Low, Increasing Medium, and Increasing High. The Stable domains are similarly labeled (SL, SM, SH). These six within-domain estimates allowed us to establish mortality and growth rates for trees of a given species that experienced similar climatic conditions. We calculated growth and mortality estimates for all CWD-based domains with 10 or more plots.

In addition to within-domain estimates, we examined growth and mortality estimate contrasts among CWD domains. This allows us to establish if trees between two CWD domains are experiencing meaningful differences in growth and mortality. For example, this allows us to ask whether trees in historically dry sites (High initial CWD) respond differently to increased water stress (Increasing, or above threshold delta CWD) compared to trees in mesic sites (Low initial CWD) experiencing similar increases, among other comparisons.

We obtained empirical means and 95% confidence intervals via a nonparametric bootstrap. For each of the 1000 bootstrap iterations we sampled FIA plots with replacement and obtained species point estimates for each domain, yielding a maximum of `r n.spp * 6 ` domain point estimates for each iteration.  For each domain we calculated a mean, upper, and lower 95% CI with the 2.5% and 97.5% quantiles from the  bootstrap iterations.  This approach allows for asymmetrical CIs around means.  We estimated contrasts between pairs of CWD domains by subtracting within-iteration domain point estimates.

Our  contrast analysis approach resulted in hundreds of comparisons. Given the large number of tests, we caution readers to interpret individual significant results (i.e., contrast 95% CIs that do not including zero) cautiously, as some proportion likely represent Type-I errors and occurred by chance.  We did not apply a multiple comparisons correction as (1) our analysis is inherently exploratory, aimed at identifying patterns within and across species rather than testing a specific *a priori* hypothesis, and (2) because we did not want to inflate the Type-II error rate and obscure real relationships in favor of reducing false positive findings.

## Dashboard
We developed a data dashboard (https://tinyurl.com/FIAgrowmort) to enable ourselves and readers to examine individual species domain estimates, the distribution of initial and delta CWD values, and the fuzzed geographic locations of plots and plot domains.  All analysis files, manuscript markdowns, dashboard code, and ancillary data sources are available at the GitHub repository https://github.com/jeremygroom/Tree_Growth_Death.

@@

Delta CWD increased at `r analysis.stats$cwd.gt0.pct`% of plots. The increases occurred in all forested regions of the three states (Figure `r fig_nums("fig_cwdmap", display = "num")`C).  Plots where CWD decreased (indicated by negative delta CWD values) are presented alone in Figure S1.  For Washington and Oregon negative CWD values were almost entirely found in the Coast Ranges and the mesic west slope of the Cascades Range crest.  The Blue Mountains and Northern and northeastern Washington lacked negative delta CWD plots entirely. In California negative delta CWD plots with values < -1.0 mm were primarily found in the northern Sierra Nevada and in the southern portions of the state. Overall, initial and delta CWD values were weakly and positively linearly associated (correlation = `r CWD.lin.cor`). The two variables were better modeled with a 3rd order polynomial although the resulting R^2^ value still indicated that the variance was not well explained by the model ($\Delta AIC$ =  `r CWD.dAIC`, R^2^ = `r round(CWD.rels$lm2$r.squared, 2)`, Figure S2, see model fits in Tables S1 and S2), implying that other processes likely affected the relationship between initial and delta CWD. For most species delta CWD neither linearly related to nor monotonically increased with initial CWD values (Figure S3; see dashboard). For instance, the relationship between the initial and delta CWD values for *Calocedrus decurrens* may be described overall with a concave curve, with some of the most negative delta CWD values occurring in the High domain (along with some of largest delta CWD values; Figure `r fig_nums("fig_domains", display = "num")`C).

@@

Initial CWD means and ranges differed among tree species (Table S3).  The `r n.spp` appeared to form two different groups by initial CWd and delta CWD (Figure `r fig_nums("fig_cwdmeans", display = "num")`).  Eight species had mean delta CWD values < 4.5 mm and mean initial CWD values < 100 mm.  *`r init.mean.sm$spp.out`* had the smallest mean value for initial CWD (`r init.mean.sm$mean.alone` mm) while *`r init.range.sm$spp.out`* had both the largest mean value (`r init.mean.lg$mean.alone` mm) and smallest range `r init.range.sm$spp.range`.  *`r init.range.lg$spp.out`* occupied the plots with the largest range of initial CWD values `r init.range.lg$spp.range` and was the most widely distributed of our species (dashboard).

@@

Overall growth rates differed by species but in some instances coincided with species’ ecological traits.  For instance, the drought-tolerant hardwoods (*Arbutus menziesii*, *Notholithocarpus densiflorus*, *Quercus chrysolepis*, and *Quercus kelloggii*) exhibited the lowest growth rate, with all mean decadal basal area increases < 75 cm^2^ (Table 1).  *Quercus chrysolepis* had the least mean growth at 40.5 cm^2^/decade.  Riparian and mesic hardwoods like *Acer macrophyllum* and *Alnus rubra* on average grew by 145 cm^2^/decade, although growth for *Acer macrophyllum* varied by state; its mean growth in California was 52.3 cm^2^/decade, less than half of its growth rate in Oregon and almost four times less than its growth rate in Washington.  High-elevation conifers like *Abies concolor* and *Picea engelmannii* would both have grown at similar rates of approximately 170-190 cm^2^/decade except that the Washington rate for *Picea engelmannii* was only 125.4 cm/decade.  The other high-elevation conifers, *Abies lasiocarpa*,  and *Tsuga mertensiana* grew more slowly at almost 70 cm^2^/decade.

@@

Individual species growth and mortality domain patterns could differ (e.g., Figures `r fig_nums("fig_domains", display = "num")`A and `r fig_nums("fig_domains", display = "num")`B). Growth domain contrasts (Figure `r fig_nums("fig_diffgrow", display = "num")`, `r t.ref("tbl_summ_grow_mort")`) did not appear to result in substantially more significant  negative or positive estimates, indicating that species overall did not exhibit more or less growth in the Increasing, Low, or High domains. The figure column “Increasing vs. Stable” displays the mean point estimate and 95% CI for contrast between corresponding Increasing and Stable domains (e.g., for *Abies lasiocarpa*, subtracting the estimates for columns SL, SM, and SH from IL, IM, IH, respectively; Figure `r fig_nums("fig_domains", display = "num")`A). Negative estimates for the “Increasing vs. Stable” in Figures `r fig_nums("fig_diffgrow", display = "num")` and `r fig_nums("fig_diffmort", display = "num")` indicate that the 10-year growth or mortality rate was greater for a species in Stable domains. `r t.ref("tbl_summ_grow_mort")` summarizes the number of domain differences in each of the columns in Figures `r fig_nums("fig_diffgrow", display = "num")` and `r fig_nums("fig_diffmort", display = "num")` where the 95% CI does not include zero as well as the number of contrasts per figure column.

We observed varying growth responses across the CWD domains (Figure `r fig_nums("fig_diffgrow", display = "num")`). Depending on the species, growth may have been susceptible (negatively or positively) to a change in CWD and/or differed by initial CWD values. We consider a lower initial CWD value to indicate a more favorable condition for plant growth, and *Abies amabilis*, *Abies grandis*, *Calocedrus decurrens*, *Pinus ponderosa*, *Pseudotsuga menziesii*, *Tsuga mertensiana*, *Acer macrophyllum*, and *Alnus rubra* exhibited greater growth in the Low domains than the Medium and/or High domains. This was exemplified by *Acer macrophyllum* and *Abies grandis* with a mean difference of `r fig.val.fcn("g", "X312", "IH - IL", 0)` cm^2^/decade, IH-IL and `r fig.val.fcn("g", "X312", "SH - SL", 0)` cm^2^/decade, SH-SL, respectively while less for contrasts for species like *Tsuga mertensiana* (approximately -0.20 for IM-IL and SM-SL) and *Quercus chrysolepis* (`r fig.val.fcn("g", "X805", "SH - SM", 0)` for SH-SM).  Except for *Alnus rubra*, these results predominantly occurred in both the Stable domain (2 spp.) or the Increasing and Stable domains (5 spp.). However, this finding is not uniform for all species, and we found that *Thuja plicata* had greater growth in the Medium domains than either High or Low domains (Increasing: `r fig.val.fcn("g", "X242", "IH - IM", 0)` cm^2^/decade for IH-IM and `r fig.val.fcn("g", "X242", "IM - IL", 0)` cm^2^/decade for IM-IL; Stable: `r fig.val.fcn("g", "X242", "SH - SM", 0)` cm^2^/decade for SH-SM and `r fig.val.fcn("g", "X242", "SM - SL", 0)` cm^2^/decade for SM-SL). *Abies concolor*, *Larix occidentalis*, *Picea engelmannii*, *Pinus contorta*, *Pinus lambertiana*, *Arbutus menziesii*, and *Quercus kelloggii* had greater growth in the more water stressed High domains than the Medium and/or Low domains, with some of the greatest differences for *Pinus lambertiana* (`r fig.val.fcn("g", "X117", "IH - IL", 0)` cm^2^/decade for IH-IL) and *Picea engelmannii* (`r fig.val.fcn("g", "X93", "SH - SL", 0)` cm^2^/decade for SH-SL).  Except for *Larix occidentalis*, these results predominantly occurred in both the Increasing domain (3 spp.) or the Increasing and Stable domains (3 spp.). Other species exhibited no significant findings (4 spp.) or displayed other result patterns (3 spp.).

@@

Regional analyses detected several species with relatively high mortality rates including *Acer macrophyllum*, *Picea engelmannii*, *Abies lasiocarpa*, *Pinus contorta*, and *Alnus rubra*.  We do not believe our results are concerning for *Alnus rubra*.  Our estimates for this short-lived tree species align with reported background mortality rates (e.g., Brown and Courtin 2018).
